D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-30 01intro_preliminary.tex
add figure

\subsection{Labelled Dirac Notation and Motivating Example}
Labelled Dirac notation is a generalization of Dirac notation for
describing multi-body quantum systems. The need for labelled Dirac
notation is motivated by the following example:
\begin{example}
  \label{example1}
  Let $p,q,r$ be three qubits and initially in the (unnormalized)
  $\ghz$ state $|\ghz\rangle \triangleq |000\rangle+|111\rangle$.
  Applying the 3-qubit Toffoli gate ($\ccnot$) with control qubits
  $p,r$ and target qubit $q$ to GHZ is equivalent to applying 2-qubit
  $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$.
  Using Dirac notation, the identity is written as:
\begin{align*}
  \ccnot|\ghz\rangle &= (I\otimes \cnot)|\ghz\rangle \\
  &= (\swap\otimes I)(I\otimes \cnot)(\swap\otimes I)|\ghz\rangle
\end{align*}
Formalizing the statement in Dirac notation requires the following
steps: 1. Arrange qubits in the conventional order, here we choose
$p,r,q$ to simplify the representation of $\ccnot$; 2. Lift the local
operation $\cnot$ to the global system. For $\cnot$ acting on $r,q$ is
easier, since $r,q$ consistent with the chosen order, and we only need
to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes
\cnot)$.  For $\cnot$ acting on $p,q$, realize that $p,q$ is not
adjacent in the chosen order, we additionally need to use $\swap$ to
temporarily exchange the qubits $p$ and $r$, i.e., globally, the
$(\swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of
$\cnot$ on $r,q$).

Roughly speaking, encoding in standard Dirac notation needs to
tensoring identity operators and employing additional $\swap$ gates
since generally the conventional order cannot guarantee 1. the order
of all local operations is consistent with it, 2. the local operations
involve only adjacent qubits in the conventional order.
\end{example}
To address the limitations of Dirac notations, physicists routinely
use labels (or subscripts) to indicate the systems on which quantum
states or operations are applied, thereby avoiding unnecessary lifting
and swap gates. For example, rewriting the above example using labels
yields:
\begin{align*}
  \ccnot_{prq}|\ghz\rangle_{pqr} = 
  \cnot_{rq}|\ghz\rangle_{pqr}
  = \cnot_{pq}|\ghz\rangle_{pqr}
\end{align*}
This formalization avoids determining and maintaining the conventional
order of qubits. nor lifting using additional $I$ and $\swap$s. In
this setting, the tensor products become an associative and
commutative (AC) symbol, so we can rearrange qubits as needed to
perform calculations. For our example, we can perform the calculation as
follows:
\begin{align*}
  &\ccnot_{prq}|\ghz\rangle_{pqr}
  = (\ccnot|000\rangle + \ccnot|111\rangle)_{prq}
  = (|000\rangle + |110\rangle)_{prq}\\
  &\cnot_{rq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{rq}|0\rangle_p + (\cnot|11\rangle)_{rq}|1\rangle_p
  =(|000\rangle + |101\rangle)_{rqp}\\
  &\cnot_{pq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{pq}|0\rangle_r + (\cnot|11\rangle)_{pq}|1\rangle_r
  = (|000\rangle + |101\rangle)_{pqr}
\end{align*}
where the RHS of each line are equivalent.  In addition, labelled
Dirac notation can conveniently describe local measurements, partial
traces (the state or evolution of subsystems in multi-body systems),
and partial inner products (the representation of partial traces in
pure states), which are sufficient to handle the mathematical formulas
of quantum mechanics in multi-body systems.

>>>>

\subsection{Labelled Dirac Notation and Motivating Example}
Labelled Dirac notation is a generalization of Dirac notation for
describing multi-body quantum systems. The need for labelled Dirac
notation is motivated by the following example:

% \begin{figure}[t]
% \begin{quantikz}
%   & \gate{X} & \ctrl{1} & \\
%   & & \targ{} &
%   \end{quantikz}
%   =\begin{quantikz}
%   & \ctrl{1} & \gate{X} & \\
%   & \targ{} & \gate{X} &
%   \end{quantikz}
%   \end{figure}

\begin{example}
  \label{example1}
  Let $p,q,r$ be three qubits and initially in the (unnormalized)
  $\ghz$ state $|\ghz\rangle \triangleq |000\rangle+|111\rangle$.
  Applying the 3-qubit Toffoli gate ($\ccnot$) with control qubits
  $p,r$ and target qubit $q$ to GHZ is equivalent to applying 2-qubit
  $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$.
  Using Dirac notation, the identity is written as:
\begin{align*}
  \ccnot|\ghz\rangle &= (I\otimes \cnot)|\ghz\rangle \\
  &= (\Swap\otimes I)(I\otimes \cnot)(\Swap\otimes I)|\ghz\rangle
\end{align*}
which might be illustrated by the following circuit models.
\begin{figure}[h]
  % \begin{minipage}{0.48\linewidth}
      \centering
      \vspace{-0.5cm}
      \begin{quantikz}[column sep=0.1em, row sep=0em]
      \lstick{$p\qquad\qquad $} & \lstick{} &[0.2cm] & \ctrl{1} & [0.2cm] &  \quad \ \ \;\! \quad\qquad\quad & 
      &[0.2cm] & & [0.2cm] & \quad \ \ \;\! \quad\qquad\quad & 
      &[0.2cm] & \gate[2,swap,style={dashed}]{} & [0.2cm] & [0.2cm]\gate[2,swap,style={dashed}]{} & [0.2cm] & \\[-0.1cm]
      \lstick{$r\qquad\qquad $} & \lstick{$|\ghz\>$} & & \ctrl{1} & & \quad = \quad\qquad\quad & \lstick{$|\ghz\>$} 
      & & \ctrl{1} & & \quad = \quad\qquad\quad & \lstick{$|\ghz\>$}
      & & & \ctrl{1} & & & \\
      \lstick{$q\qquad\qquad $} & \lstick{} & & \targ{} & & \quad \ \ \;\! \quad\qquad\quad & 
      & & \targ{} & & \quad \ \ \;\! \quad\qquad\quad &
      & & & \targ{} & & & \\
      \end{quantikz}
      \vspace{-0.5cm}
      % \vspace{-0.7em}
      % \caption{Circuit computing the last bit of $a+(011)_2$.}
      % \label{fig:addition}
  %  \end{minipage}
  \end{figure}
Formalizing the statement in Dirac notation requires the following
steps: 1. Arrange qubits in the conventional order, here we choose
$p,r,q$ to simplify the representation of $\ccnot$; 2. Lift the local
operation $\cnot$ to the global system. For $\cnot$ acting on $r,q$ is
easier, since $r,q$ consistent with the chosen order, and we only need
to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes
\cnot)$.  For $\cnot$ acting on $p,q$, realize that $p,q$ is not
adjacent in the chosen order, we additionally need to use $\Swap$ to
temporarily exchange the qubits $p$ and $r$, i.e., globally, the
$(\Swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of
$\cnot$ on $r,q$).

Roughly speaking, encoding in standard Dirac notation needs to
tensoring identity operators and employing additional $\Swap$ gates
since generally the conventional order cannot guarantee 1. the order
of all local operations is consistent with it, 2. the local operations
involve only adjacent qubits in the conventional order.
\end{example}
To address the limitations of Dirac notations, physicists routinely
use labels (or subscripts) to indicate the systems on which quantum
states or operations are applied, thereby avoiding unnecessary lifting
and swap gates. For example, rewriting the above example using labels
yields:
\begin{align*}
  \ccnot_{prq}|\ghz\rangle_{pqr} = 
  \cnot_{rq}|\ghz\rangle_{pqr}
  = \cnot_{pq}|\ghz\rangle_{pqr}
\end{align*}
This formalization avoids determining and maintaining the conventional
order of qubits, nor lifting using additional $I$ and $\Swap$s. In
this setting, the tensor products become an associative and
commutative (AC) symbol, so we can rearrange qubits as needed to
perform calculations. For our example, we can perform the calculation as
follows:
\begin{align*}
  &\ccnot_{prq}|\ghz\rangle_{pqr}
  = (\ccnot|000\rangle + \ccnot|111\rangle)_{prq}
  = (|000\rangle + |110\rangle)_{prq}\\
  &\cnot_{rq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{rq}|0\rangle_p + (\cnot|11\rangle)_{rq}|1\rangle_p
  =(|000\rangle + |101\rangle)_{rqp}\\
  &\cnot_{pq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{pq}|0\rangle_r + (\cnot|11\rangle)_{pq}|1\rangle_r
  = (|000\rangle + |101\rangle)_{pqr}
\end{align*}
where the RHS of each line are equivalent.  In addition, labelled
Dirac notation can conveniently describe local measurements, partial
traces (the state or evolution of subsystems in multi-body systems),
and partial inner products (the representation of partial traces in
pure states), which are sufficient to handle the mathematical formulas
of quantum mechanics in multi-body systems.

### 2025-01-30 03labelled.tex
add figure

\begin{definition}[Cylindrical Extension]
  For any $D : \cD(s_1,s_2)$ and $s$ that disjoint with both $s_1$ and $s_2$, we define $\cl(D,s) \triangleq D\otimes \mathbf{1}_{s}$ of type $\cD(s_1\cup s, s_2\cup s)$.
\end{definition}
Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names. For any valid register $R$, there exists the operator $\swap_R$ that sorts $R$; for example, if $E\vdash p : \reg(\sigma_p)$ and $E\vdash q : \reg(\sigma_q)$, then $\swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$. We can further define $\swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.

For any interpretation $\sem{\cdot}$, and write $\cH_{\sem{s}} \triangleq \bigotimes_{r\in s}\cH_{\sem{\sigma_r}}$ for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. We interpret labelled Dirac notations inductively as (see details in Appendix) : assume $E\vdash D_i : \cD(s_i,s'_i)$ in the following,
\begin{itemize}
  \item $\sem{|i\>_r} = |\sem{i}\>$; $\sem{K_R} = \swap_R \cdot \sem{K}$; $\sem{O_{R_1,R_2}} = \swap_{R_1}\cdot \sem{O}\cdot \swap_{R_2}$;
  \item $\sem{D_1\otimes \cdots \otimes D_n} = \swap_{s_1,\cdot,s_n}\cdot (\sem{D_1}\otimes \cdots \otimes \sem{D_1})\cdot \swap_{s'_1,\cdot,s'_n}$;
  \item $\sem{D_1\cdot D_2} = \sem{\cl(D_1, s_2\backslash s'_1)}\cdot \sem{\cl(D_2, s'_1\backslash s_2)}$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
Thus, $\sem{D}$ is a linear map from $\cH_{\sem{s'}}$ to $\cH_{\sem{s}}$ given $E\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2} = \sem{D_2\otimes D_1}$ and $\sem{D_1\otimes (D_2\otimes D_3)} = \sem{(D_1\otimes D_2)\otimes D_3}$, which ensures the soundness of treating labelled tensor as an AC symbol.

>>>>

\begin{definition}[Cylindrical Extension]
  For any $D : \cD(s_1,s_2)$ and $s$ that disjoint with both $s_1$ and $s_2$, we define $\cl(D,s) \triangleq D\otimes \mathbf{1}_{s}$ of type $\cD(s_1\cup s, s_2\cup s)$.
\end{definition}
Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names. For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $E\vdash p : \reg(\sigma_p)$ and $E\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$. We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.

For any interpretation $\sem{\cdot}$, and write $\cH_{\sem{s}} \triangleq \bigotimes_{r\in s}\cH_{\sem{\sigma_r}}$ for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. We interpret labelled Dirac notations inductively as (see details in Appendix) : assume $E\vdash D_i : \cD(s_i,s'_i)$ in the following,
\begin{itemize}
  \item $\sem{|i\>_r} = |\sem{i}\>$; $\sem{K_R} = \Swap_R \cdot \sem{K}$; $\sem{O_{R_1,R_2}} = \Swap_{R_1}\cdot \sem{O}\cdot \Swap_{R_2}$;
  \item $\sem{D_1\otimes \cdots \otimes D_n} = \Swap_{s_1,\cdot,s_n}\cdot (\sem{D_1}\otimes \cdots \otimes \sem{D_1})\cdot \Swap_{s'_1,\cdot,s'_n}$;
  \item $\sem{D_1\cdot D_2} = \sem{\cl(D_1, s_2\backslash s'_1)}\cdot \sem{\cl(D_2, s'_1\backslash s_2)}$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
Thus, $\sem{D}$ is a linear map from $\cH_{\sem{s'}}$ to $\cH_{\sem{s}}$ given $E\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2} = \sem{D_2\otimes D_1}$ and $\sem{D_1\otimes (D_2\otimes D_3)} = \sem{(D_1\otimes D_2)\otimes D_3}$, which ensures the soundness of treating labelled tensor as an AC symbol.
